TrenchR: an R package for transparent environmental and ecological biophysics

### 2022-06-28 Manuscript_full.Rmd
supp rendering and bib
more bib
adding the tex preamble for the listings package following the rmarkdown cookbook, which controls formatting of the code blocks

---
title: 'TrenchR: an R package for transparent environmental and ecological biophysics'
author: "Lauren B. Buckley, Bryan A. Briones Ortiz, Isaac Caruso, Aji John, Ofir Levy, Abigail V. Meyer, Eric A. Riddell, Yutaro Sakairi, and Juniper Simonis"
output:
  html_document: default
  word_document: default
  pdf_document: default
bibliography: bibliography.bib
csl: methods-in-ecology-and-evolution.csl
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

@@

The mathematical tools for leveraging environmental data to predict organismal conditions have long lingered in books [@campbell2000introduction; @gates1980biophysical] and articles [@porter1969thermodynamic; @porter1983biophysical], but adoption of these tools has not kept pace with research on climate change responses.
In recent decades, adoption of biophysical approaches is expanding with recognition of the importance of spatial and temporal environmental variation to determining climate change responses [@buckley2018leveraging; @kearney2017nichemapr] and the increased dissemination of computational tools.

Estimating how organisms experience their thermal environment generally entails two classes of models [@campbell2000introduction; @gates1980biophysical] (Figure 1).
Microclimate models allow scaling conditions from sensors to organism height. They can characterize heat and air transport to estimate vertical air and soil temperature and wind profiles. 
They can also characterize direct and diffuse solar radiation and longwave radiation emanating from the sky, ground, and surrounding objects like vegetation. 
Biophysical models estimate body temperatures using an energy budget to balance heat exchanges between organisms and their environment.

@@

For additional background on microclimate and biophysical modeling, we have updated a series of tutorials entitled *Physical Processes in Ecosystems*. 
We intend the tutorials to provide a more contained and accessible introduction to microclimate and biophysical modeling than that included in classic textbooks [@campbell2000introduction; @gates1980biophysical]. 
The tutorials provide less detailed coverage of many of the topics contained in the books. 
The series of 14 tutorials starts with an overview of the calculus and physics principles underlying the modeling. 
Thermodynamics and energy budget modeling are then introduced with detailed examples. 
Tutorials address the climate space concept, operative temperatures, biophysical models for leaves and sheep, and heat flow in soils. 
We have revised the tutorials (which originated from an NSF training grant in 1979 and lacked broad distribution) to include R code and utilization of TrenchR functions. 
We expanded the original series to include tutorials contributed by M. Kearney introducing the Microclim environmental data [@kearney2014microclim] and integrating the Microclim data with biophysical modeling to examine broad scale climatic limits. 
The tutorials are provided in html form and available for download as a pdf at the bookdown server (<https://bookdown.org/huckley/Physical_Processes_In_Ecosystems/>).
R markdown files for the tutorials are available in Github (<https://github.com/trenchproject/TrenchRmodules>).

@@

We illustrate the use of the TrenchR package by estimating an energy budget for a *Sceloporus* lizard on June 1, 2021 in Santa Fe, New Mexico, USA (35.69$^\circ$, -105.944$^\circ$, elevation: 2121 m). 
The example is also incorporated in the *Using energy balances to estimate body temperatures* vignette. 
We start by generating environmental inputs (Figure 2).
Using these inputs, we estimate the energy budget with component functions. 
Finally, we use an integrated biophysical model to estimate operative environmental temperatures. 
See Appendix S1 for another example examining a time series of microclimate data and applying and testing the biophysical models.

@@

Let us assume the lizard is in an unshaded location where a weather station at standard height (2 meters) reports that the daily air temperature varies from a minimum of 10$^\circ$C to a maximum of 25$^\circ$C and the wind speed averages 1 m/s. 
The soil surface temperature varies from a minimum of 15$^\circ$C to a maximum of 30$^\circ$C. 
We assume that atmospheric transmissivity $\tau = 0.7$ and albedo $\rho = 0.6$.

@@

```{r, echo=FALSE}
library(TrenchR)
```

>>>>

---
title: 'TrenchR: an R package for transparent environmental and ecological biophysics'
author: 
  - "Lauren B. Buckley, Bryan A. Briones Ortiz, Isaac Caruso, Aji John, Ofir Levy,"
  - "Abigail V. Meyer, Eric A. Riddell, Yutaro Sakairi, and Juniper Simonis"
output:
  html_document: default
  word_document: default
  pdf_document: 
    pandoc_args: --listings
    includes:
      in_header: preamble.tex
bibliography: bibliography.bib
csl: methods-in-ecology-and-evolution.csl
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(TrenchR)

@@

The mathematical tools for leveraging environmental data to predict organismal conditions have long lingered in books [@gates1980biophysical; @campbell2000introduction] and articles [@porter1969thermodynamic; @porter1983biophysical], but adoption of these tools has not kept pace with research on climate change responses.
In recent decades, adoption of biophysical approaches is expanding with recognition of the importance of spatial and temporal environmental variation to determining climate change responses [@buckley2018leveraging; @kearney2017nichemapr] and the increased dissemination of computational tools.

Estimating how organisms experience their thermal environment generally entails two classes of models [@gates1980biophysical; @campbell2000introduction] (Figure 1).
Microclimate models allow scaling conditions from sensors to organism height. They can characterize heat and air transport to estimate vertical air and soil temperature and wind profiles. 
They can also characterize direct and diffuse solar radiation and longwave radiation emanating from the sky, ground, and surrounding objects like vegetation. 
Biophysical models estimate body temperatures using an energy budget to balance heat exchanges between organisms and their environment.

@@

For additional background on microclimate and biophysical modeling, we have updated a series of tutorials entitled *Physical Processes in Ecosystems*. 
We intend the tutorials to provide a more contained and accessible introduction to microclimate and biophysical modeling than that included in classic textbooks [@gates1980biophysical; @campbell2000introduction]. 
The tutorials provide less detailed coverage of many of the topics contained in the books. 
The series of 14 tutorials starts with an overview of the calculus and physics principles underlying the modeling. 
Thermodynamics and energy budget modeling are then introduced with detailed examples. 
Tutorials address the climate space concept, operative temperatures, biophysical models for leaves and sheep, and heat flow in soils. 
We have revised the tutorials (which originated from an NSF training grant in 1979 and lacked broad distribution) to include R code and utilization of TrenchR functions. 
We expanded the original series to include tutorials contributed by M. Kearney introducing the Microclim environmental data [@kearney2014microclim] and integrating the Microclim data with biophysical modeling to examine broad scale climatic limits. 
The tutorials are provided in html form and available for download as a pdf at the bookdown server (<https://bookdown.org/huckley/Physical_Processes_In_Ecosystems/>).
R markdown files for the tutorials are available in Github (<https://github.com/trenchproject/TrenchRmodules>).

@@

We illustrate the use of the TrenchR package by estimating an energy budget for a *Sceloporus* lizard on June 1, 2021 in Santa Fe, New Mexico, USA (35.69$^\circ$N, -105.944$^\circ$W, elevation: 2121 m). 
The example is also incorporated in the *Using energy balances to estimate body temperatures* vignette. 
We start by generating environmental inputs (Figure 2).
Using these inputs, we estimate the energy budget with component functions. 
Finally, we use an integrated biophysical model to estimate operative environmental temperatures. 
See Appendix S1 for another example examining a time series of microclimate data and applying and testing the biophysical models.

@@

Let us assume the lizard is in an unshaded location where a weather station at standard height (2 meters) reports that the daily air temperature varies from a minimum of 10 $^\circ$C to a maximum of 25 $^\circ$C and the wind speed averages 1 m/s. 
The soil surface temperature varies from a minimum of 15 $^\circ$C to a maximum of 30 $^\circ$C. 
We assume that atmospheric transmissivity $\tau = 0.7$ and albedo $\rho = 0.6$.

@@
